---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ESPENAPI

<!-- badges: start -->
<!-- badges: end -->

The goal of ESPENAPI is to provide access to Neglected Tropical Diseases (NTDs) data from the ESPEN portal and provide a minimum visualisation/summary of the data. Instead of having to visit the ESPEN portal to download the data, this R package allows the user to acquire the data and process it directly from R. A web application to visualise the data is currently under development, a preliminary version can be found here <https://olatunjijohnson.shinyapps.io/espenshiny/>

## Installation

<!-- You can install the released version of ESPENAPI from [CRAN](https://CRAN.R-project.org) with:

``` r
install.packages("ESPENAPI")
```
-->

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("olatunjijohnson/ESPENAPI", ref="main")
```
## Example

This is a basic example which shows how to download the STH data at site level from Kenya for 2010.

```{r example}
library(ESPENAPI)
data  <- ESPENAPI::ESPEN_API_data(country="Kenya", disease="sth", level="sitelevel", start_year=2010, end_year=2010)
## basic example code
```

## Further improvements

1. Allows user to only supply api_key once and it saves into the environment 

2. Develop another function to download maps from ESPEN portal

3. Develop dashboard to summarise the data
