---
title: "Introduction to ESPENAPI"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ESPENAPI}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE
)
```

## What is ESPEN?

The **Expanded Special Project for Elimination of Neglected Tropical
Diseases** (ESPEN) is a WHO initiative that coordinates the surveillance
and control of NTDs in the WHO African Region. The ESPEN portal hosts
country-level and site-level programme data for seven diseases:

| Code | Disease |
|------|---------|
| `lf` | Lymphatic filariasis |
| `oncho` | Onchocerciasis (river blindness) |
| `loa` | Loiasis (eye worm) |
| `sch` | Schistosomiasis (bilharzia) |
| `sth` | Soil-transmitted helminths |
| `trachoma` | Trachoma |
| `coendemicity` | Co-endemicity reporting |

The ESPENAPI package provides direct R access to this data through
the ESPEN portal API, so you can download, analyse, and visualise
NTD data entirely within R.

---

## Getting started

### Step 1: Get an API key

Request a free API key from the ESPEN portal:
<https://espen.afro.who.int>

### Step 2: Store the key safely

Never paste your key directly into scripts — store it in your `.Renviron`
file instead:

```{r key-setup}
# Run this once to open your .Renviron file
usethis::edit_r_environ()
```

Add this line, replacing the placeholder:

```
ESPEN_API_KEY=your_key_here
```

Save the file and restart R. The key will then be found automatically
by all ESPENAPI functions.

For guided instructions, call:

```{r key-setup2}
espen_key_setup()
```

---

## Basic usage

### Download data for a single country

```{r single-country}
library(ESPENAPI)

# STH data for Nigeria at site level, 2010–2015
dat <- ESPEN_API_data(
  country    = "Nigeria",
  disease    = "sth",
  level      = "sitelevel",
  start_year = 2010,
  end_year   = 2015
)

dim(dat)
head(dat)
```

### Explore what's available

```{r reference-tables}
# All valid disease codes
espen_diseases()

# All valid spatial levels
espen_levels()
```

---

## Choosing the spatial level

ESPEN data is available at two spatial levels:

**Implementation unit (`level = "iu"`)** is the administrative unit used
for MDA (mass drug administration) programme planning — typically a
district or sub-district. Use this for programme coverage and
endemicity summaries.

```{r iu-level}
lf_iu <- ESPEN_API_data(
  country    = "Ghana",
  disease    = "lf",
  level      = "iu",
  start_year = 2015,
  end_year   = 2020
)
head(lf_iu)
```

**Site level (`level = "sitelevel"`)** contains individual survey or
sentinel site results, including sample sizes and prevalence estimates.
Use this for epidemiological analysis.

```{r site-level}
lf_sites <- ESPEN_API_data(
  country    = "Ghana",
  disease    = "lf",
  level      = "sitelevel",
  start_year = 2015,
  end_year   = 2020
)
head(lf_sites)
```

---

## Downloading data for multiple countries

Pass a vector of country names (or ISO2 codes) to get data for all of
them in one combined data frame:

```{r multi-country}
# Three countries in one call
multi <- ESPEN_API_data(
  country    = c("Nigeria", "Ghana", "Kenya"),
  disease    = "sth",
  level      = "sitelevel",
  start_year = 2010,
  end_year   = 2015,
  verbose    = TRUE   # shows progress for each country
)

nrow(multi)
table(multi$Country)
```

ISO2 codes work the same way:

```{r iso2-multi}
multi_iso <- ESPEN_API_data(
  iso2       = c("NG", "GH", "KE"),
  disease    = "sth",
  level      = "sitelevel",
  start_year = 2010,
  end_year   = 2015
)
```

---

## Forecast data (MDA and impact assessment)

To request forecast data rather than observed data, set `type = TRUE`
and specify a `subtype`:

```{r forecast}
# MDA forecast for Nigeria
mda <- ESPEN_API_data(
  country = "Nigeria",
  disease = "sth",
  level   = "iu",
  type    = TRUE,
  subtype = "mda"
)

# Impact assessment forecast
impact <- ESPEN_API_data(
  country = "Nigeria",
  disease = "sth",
  level   = "iu",
  type    = TRUE,
  subtype = "impact_assessment"
)
```

---

## Limiting columns returned

Use `attributes` to request only specific columns — useful for large
downloads:

```{r attributes}
slim <- ESPEN_API_data(
  country    = "Nigeria",
  disease    = "sth",
  level      = "iu",
  start_year = 2015,
  end_year   = 2020,
  attributes = "IU_ID,Endemicity,MDA,EffMDA"
)
names(slim)
```

---

## Pagination

For very large queries, use `limit` and `offset` to page through results:

```{r pagination}
# First 500 records
page1 <- ESPEN_API_data(
  country = "Nigeria", disease = "sth",
  level   = "sitelevel",
  limit   = 500, offset = 0
)

# Next 500 records
page2 <- ESPEN_API_data(
  country = "Nigeria", disease = "sth",
  level   = "sitelevel",
  limit   = 500, offset = 500
)
```

---

## Accessing the raw JSON

Set `df = FALSE` to get the raw API response as an `ESPEN_api` object
rather than a parsed data frame. This is useful for debugging or when
you want to handle the JSON parsing yourself.

```{r raw}
raw <- ESPEN_API_data(
  country = "Nigeria",
  disease = "sth",
  level   = "sitelevel",
  df      = FALSE
)

# Print method shows status and size
print(raw)

# Access content
cat(substr(raw$content, 1, 200))
```

---

## Worked example: comparing STH prevalence across West Africa

```{r worked-example}
library(ESPENAPI)
library(ggplot2)

# Download site-level STH data for West African countries
west_africa <- ESPEN_API_data(
  country    = c("Nigeria", "Ghana", "Senegal", "Mali", "Burkina Faso"),
  disease    = "sth",
  level      = "sitelevel",
  start_year = 2010,
  end_year   = 2020,
  verbose    = TRUE
)

# Example: plot record counts per country per year
library(dplyr)

west_africa %>%
  group_by(Country, Year) %>%
  summarise(n = n(), .groups = "drop") %>%
  ggplot(aes(x = Year, y = n, colour = Country)) +
  geom_line() +
  geom_point() +
  labs(
    title = "ESPEN STH site-level records",
    y     = "Number of survey sites",
    x     = "Year"
  ) +
  theme_bw()
```
